### 2024-05-14 Paper/paper.tex
Updates from Overleaf

\usepackage{tgtermes}
\usepackage{epsf}
\usepackage{epstopdf}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage[colorlinks=true,linkcolor=blue,citecolor=blue]{hyperref}
\usepackage{dcolumn}
\usepackage{amsmath, amsthm, amssymb}
\usepackage{mwe}
\usepackage{url}
%\usepackage{harvard}
\usepackage{fancyheadings}
\usepackage{longtable}
\usepackage{authblk}
\usepackage{setspace}
%\usepackage[nomarkers]{endfloat}
\usepackage{float}
\usepackage{bbm}
%\usepackage{titling}
\usepackage{subcaption}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{import}
\usepackage[backend=biber,style=authoryear,
sorting=ynt,citestyle=authoryear]{biblatex}
\addbibresource{papercitations.bib}
%\usepackage[nomarkers,nofiglist,notablist]{endfloat}

@@

    \begin{abstract}
		{\small
        In many industries, not-for-profit (NFP) make up the majority of firms in the market. Understanding the objectives and behaviors of these firms is important when creating incentives. However, NFP firms have goals and missions that make understanding their objectives difficult. For hospitals in particular, there is mixed evidence on whether NFP hospitals differ meaningfully from for-profit (FP) hospitals. In this paper, I investigate whether leadership team composition of nonprofit hospitals affects how similarly they behave to for-profits. I use a novel data set on NFP hospital executives in the US (2009-2015) to investigate whether the presence of clinically trained executives affects hospital behavior. In particular, I leverage two pay-for-performance initiatives in the US enacted in 2012, and estimate the difference in readmission and mortality rates afterwards between FP, NFP, and NFP with and without clinically trained executives. I find that NFPs without clinically trained executives behave more like FPs than other NFPs. This is consistent with theoretical predictions under the assumption that NFP hospitals with clinically trained executives value societal outcomes more than those without clinically trained executives.
		} 
	\end{abstract}

@@

    This paper contributes to three strands of literature. First, it adds to our understanding of what underlies NFP hospital behavior. There is mixed evidence on whether NFP behavior differs from FP behavior: costs, uncompensated care, technology adoption, and quality (\cite{sloan2000not}; \cite{eggleston2008hospital}; \cite{moscelli2018effect}; \cite{moscone2020public}). Underlying hospital are hospital objectives. To this point, researchers have found that NFPs do not act as purely profit maximizers, but maximize some combination of profit and output (\cite{deneffe2002not}, \cite{chang2011nonprofit}). A paper with an approach similar to mine which investigates hospital behavior is \citeauthor{chang2011nonprofit} (\citeyear{chang2011nonprofit}), which looks at the effect of a cost shock on likelihood of shutting down and mix of profitable services. They find that NFPs and FPs are equally as likely to shut down after the shock, but only NFPs adjust their mix of profitable services. Similarly, I investigate a change in incentives to hospitals, but I look at the differential impacts of leadership teams within nonprofits. This paper adds to our understanding of inputs into nonprofit hospital objective functions.

@@

    \subsection{Nonprofit Hospital Leadership}

    NFPs are governed by a board of directors, whose role is to set broad goals and strategies, and provide general oversight. The board selects executive team members to carry out day-to-day operations. While hospital executives typically specialize by getting degree(s) in healthcare management or an MBA specific to healthcare, they do not often have clinical training. While some doctors earn additional degrees before stepping into an executive role, this is not a necessary condition to becoming a physician executive. An executive team usually consists of at least a CEO and CFO, but every NFP is different in exactly how they structure their executive team. Some NFPs hire Chief Medical Officers, Chief Quality Officers, executive directors, vice presidents of various departments, etc. I consider all of these employees as part of the day-to-day management of the hospital and include them as a part of the executive team.

    All tax-exempt organizations in the US are required to file a Form 990 with Internal Revenue Services (IRS) each year. There are different types of forms, but any organization grossing over \$200,000 must file the most extensive Form 990. Sections of this form include a statement of revenue, statement of functional expenses, a balance sheet, and, as used in this project, a list of all key employees, executives, and board members. In this section, each firm is required to report the name and title, average hours per week, position, and compensation of their board members and executive leaders. Since these firms are tax exempt, their information in these forms is made publicly available.

    \subsection{Pay-for-Performance Programs}\label{sec:hrrp}

    The Hospital Readmissions Reduction Program (HRRP) was passed as a part of the Affordable Care Act in 2010 as a pay-for-performance incentive for hospitals to increase quality. In October 2011, the Center for Medicare and Medicaid Services (CMS) released a set of rules that mandated penalties for hospitals with above average readmission rates. A readmission is when a patient returns to the hospital within 30 days of being discharged from a previous stay; avoidable readmissions are a bad outcome for patients and increase health care spending. The goal of HRRP is to lower readmissions through better care coordination, less initial stay complications, and better post-care instructions. Beginning in October 2012, hospitals with higher readmission rates than the national average in pneumonia, heart failure, or AMI (after adjusting for demographic characteristics) receive a lower reimbursement rate for Medicare patients. In 2015, CMS also included chronic obstructive pulmonary disease, coronary artery bypass graft surgery, and elective primary total hip arthroplasty and/or total knee arthroplasty as conditions which go into the penalty calculation. \textcolor{red}{(add citations??)}

    Penalized hospitals pay penalties in the form of a fixed rate for every Medicare patient regardless of the condition. Further, CMS does not distinguish a necessary readmission from an avoidable readmission; any repeat hospital visit is included in the penalty calculation. Excess readmission rates are calculated using a rolling look-back period of 3 years to determine whether the hospital is penalized. Therefore, hospitals had incentive to react immediately once details of the program were announced in October of 2011. These penalties are not insignificant; penalized hospitals paid, on average, 4-5\% of revenue.

    \textcolor{red}{(is the program still going?)}

@@

    \subsection{Names of Hospital Leaders}

    To characterize hospitals as having clinically trained executives or not, I construct a novel data set of nonprofit hospitals in the US that contains names and select characteristics of all executives tied to the hospital from 2009-2015. I gather this data from publicly available Tax Form 990s, which every sufficiently large nonprofit files each year in the US. To my knowledge, this is the first large scale gathering of executive names from these forms.

    The historical tax forms for all nonprofits are housed by ProPublica.\footnote{https://projects.propublica.org/nonprofits/} I use the NonProfit Explorer API to extract the Employee Identification Numbers (EIN) of all nonprofits classified as a hospital according to their National Taxonomy of Exempt Entities (NTEE) code. After filtering out associations and specialty hospitals, there are around 3,000 firms left. Importantly, I need to be able to link hospitals in the tax forms to other hospital characteristics from publicly available sources. No official crosswalk between EINs and AHA IDs or Medicare numbers exists, so I must rely on matching by location and name of hospital in both data sets. After limiting to only general, nonprofit hospitals that I can link to the AHA survey, I confidently match approximately 1200 EINs to an AHA identification number based on exact name matches within the same state.

    For each of the 1200 EINs, I then extract web URLs to Form 990 PDFs in each year. I download these locally and use optical character recognition (OCR) text extraction methods to scrape the relevant section of the pdf, which is titled ``Officers, Directors, Trustees, Key Employees, and Highest Compensated Employees". Finally, I use string cleaning methods to identify names and positions for each EIN, year. I also identify titles indicating a physician: MD, Dr., or DO. Using OCR extraction can be unreliable in some instances when the text is handwritten or the font is too small. Therefore, I manually find names, positions, and titles for observations that are missing after the initial extraction. This yields 850 EINs that are matched to AHA data, and are not missing text information.

    Due to the lack of consistency in organization of executive teams, I focus mainly on the executive teams as a whole instead of specific positions. I drop anyone who is solely a board member, and focus on those who have executive in their title or are labeled as a president or vice president of a specific aspect of the hospital. I create various hospital-level characterizations based on their executive teams, such as the existence of a clinically trained executive, the number of clinically trained executives, the number of total executives, and whether the hospital has a Chief Medical Officer.

    \subsection{Other Hospital Characteristics and Outcomes}

    I include hospital information from the AHA survey on the number of beds in the hospital and the Medicare number, which I use to merge the Hospital Compare and Healthcare Cost Reporting Information System (HCRIS) data. Further, I use the AHA categorization of for-profit or nonprofit. Hospital Compare contains information on penalties from HRRP and readmission and mortality rates for the relevant HRRP conditions: heart failure, heart attack (AMI), and pneumonia. From HCRIS, I gather information on penalties from the HVBP program, operating expenses, and uncompensated care. I also link to the publicly available CMS file containing each hospital's case mix index.

    The Hospital Compare excess readmissions files contain information on the excess readmission measure used by CMS to determine whether a hospital should be penalized. Specifically, CMS calculates the average readmission rate of all hospitals, adjusts for demographic factors, and determines the amount of readmissions a hospital has in excess to of the average. This variable is condition specific, where a hospital is penalized if any of the three excess readmission ratios are greater than one\footnote{Additional conditions were added to the penalty calculations in 2015, but I do not consider these since my sample does not extend past 2015}. Further, Hospital Compare also publishes hospital level readmission and mortality rates for heart failure, AMI, and pneumonia. For my main outcomes of interest, I take an average across conditions, weighted by the total number of patients seen for that condition.

    \subsection{Summary Statistics}

    \import{Tables}{overall_sumstats.tex}

@@

    \import{Tables}{stable_sample_sumstats.tex}

    In Table \ref{sumstats}, I present summary statistics for hospital variables. There are 3,766 total hospitals in the sample, with 150 beds on average. Hospitals in the sample are less likely to be penalized for AMI than either heart failure or pneumonia under the HRRP program. On average, 20\% of patients with one of these conditions would be readmitted. While readmission rates for the specific conditions are similar, the highest likelihood of readmission is for heart failure patients. Alternatively, AMI is the most dangerous condition in terms of mortality rates.

    I present a table of means for each hospital type (for-profit, all nonprofits, nonprofits with a clinically trained executive, and nonprofits without a clinically trained executive) in Table \ref{tab:sumstats_samples}. Eighty-five percent of hospital in the sample are classified as nonprofits. Importantly, the two sub-samples of nonprofits do not add up to the total number of nonprofits because of the limited ability to extract leadership text for all nonprofits. There are key differences in nonprofits with and without clinically trained executives: those with MDs are larger, more likely to be penalized, have higher uncompensated care, and more complex patients.

    %\import{Tables}{sample_sumstats.tex}

    I plot readmission and mortality rates across time for different hospital types in Figure \ref{fig:weighted_read_mort_graph}. While this is purely descriptive evidence, the trends in outcomes point to more similar behavior between for-profits and nonprofits without clinically trained executives in comparison to nonprofits with a clinically trained executive.

    \begin{figure}[ht!]
    \centering
        \caption{Outcomes over time in all penalized hospitals}
        \includegraphics[width=\textwidth]{Objects/weighted_read_mort_graph.pdf}
        \label{fig:weighted_read_mort_graph}
    \end{figure}

    \subsection{Leadership Correlation with Other Hospital Characteristics}

    Finally, I present a table of correlations between having at least one clinically trained executive and other time invariant hospital characteristics. There may be a concern that this particular characteristic of executive teams is simply a proxy for another quality, such as being an academic medical center. Unsurprisingly, the most correlated characteristic with having a clinically trained executive is having a CMO on the executive team. However, even this correlation is only .4, meaning there is still variation in having clinically trained executives even when controlling for having a CMO.

    \import{Tables}{hosp_correlations.tex}

@@

    I model hospital behavior as maximizing some weighted average of profit and societal (or patient) benefit. Abstracting away from quantity and price as functions of quality for the purpose of this paper, I present a two-period model of hospital behavior that captures the change in quality that occurs after pay-for-performance incentives are introduced.

    \subsection{Pre-Program Incentives}

    In the first period, prior to program enactment, hospitals do not expect any penalties in the future, and choose quality according to:

    \noindent where $R$ is net revenue, $c_{\theta}(.)$ is an increasing cost function specific to quality $\theta$, $u(.)$ is extra utility gained from quality, which is increasing and concave is $\theta$, and $\alpha\in[0,1]$ captures how much the hospital cares about profit vs. societal benefit. There is an implicit stay open condition throughout the model. Taking the first order condition yields

    $$\frac{du'(\theta)}{d\alpha} = c_{\theta}'(\theta)\frac{1}{(1-\alpha)^2} > 0.$$

    \noindent That is, the more weight the hospital places on profit, the lower quality of care they will provide.

    \subsection{Post-Program Incentives}

    For simplicity, I model pay-for-performance incentives as a simplified version of how they were applied in HRRP, as penalties for low quality hospitals. However, these results still hold when including incentive payments as in HVBP.

    Hospitals determine period 1 quality, $\theta_1$, according to the optimal $\theta(\alpha)$ described above. Then, in period two, hospitals are penalized or not depending on $\theta_1$, and they choose a new quality, $\theta_2$, which will determine penalties in the future. The penalty $(k)$ comes in the form of a negative shock to revenue for hospitals with quality below a certain exogenous threshold, $\overline{\theta}$. That is, new revenue is:

    $$R_2 = \begin{cases}
        R - k, \text{  if  } \theta_1<\overline{\theta}\\
        R, \text{  if  } \theta_1\geq \overline{\theta}.
    \end{cases}$$

    The penalty $k$ is large and exceeds the maximum cost of quality. Therefore, all hospitals will choose $\theta_2$ to not be penalized in the future. Hospitals who were penalized will increase quality just to the point of not being penalized, and hospitals who were not penalized will not change their quality. Thus,

    $$\theta_2 = \begin{cases}
        \overline{\theta}, \text{  if  } \theta_1<\overline{\theta}\\
        \theta_1, \text{  if  } \theta_1\geq \overline{\theta}.
    \end{cases}$$

    \noindent Taking the change in theta from period 1 to period 2, and then differentiating with respect to $\alpha$, yields:

    $$\Delta\theta = \theta_2-\theta_1 = \begin{cases}
        \overline{\theta}-\theta_1, \text{  if  } \theta_1<\overline{\theta}\\
        0, \text{  if  } \theta_1\geq \overline{\theta}.
    \end{cases}$$

    $$\frac{d\Delta\theta}{d\alpha} = \begin{cases}
        -\frac{d\theta_1}{d\alpha}, \text{  if  } \theta_1<\overline{\theta}\\
        0, \text{  if  } \theta_1\geq \overline{\theta}.
    \end{cases}$$

    I show in period one that $\frac{d\theta}{d\alpha}<0$. Therefore, $\frac{d\Delta\theta}{d\alpha}\geq0$. That is, the response to the program incentives is weakly increasing in $\alpha$, and there is a threshold for which increasing $\alpha$ (a hospital with more weight on profit), increases the response to the penalty. Intuitively, the programs change hospital incentives in that quality now enters directly into the profit function. Therefore, hospitals who place more weight on profit will have a greater response to the program incentives.

    The theoretical prediction is that change in quality depends on how much weight the hospital places on profit vs. societal benefit. However, $\alpha$s are not observed in reality. Thus, I empirically investigate whether the response to the program is correlated with hospital ownership or clinical training on the executive team in order to gain understanding about how these things point to an underlying $\alpha$.

    First, I test empirically whether different sub-samples of nonprofits respond to program incentives differently than for-profits. While nonprofit hospital objectives are largely unknown, for-profits have revealed their objective in their ownership classification. That is, $\alpha=1$ (or is very close). The goal in this analysis is to learn something about the underlying $\alpha$ for nonprofit firms, and determine if executive team composition is a characteristic that points to differing $\alpha$s even among nonprofits. Thus, I compare their program response to that of for-profits. I estimate the following regression equation:

    \noindent where $y_{ht}$ is one of the outcome variables discussed in Section \ref{sec:data}, $\beta$ is the coefficient of interest capturing the effect of being for-profit after the programs were introduced, and $\gamma_h$ and $\delta_t$ are hospital and time fixed effects, respectively. For each outcome, I estimate three models where the comparison group differs by hospital type: all nonprofits, nonprofits with a clinically trained executive, and nonprofits without a clinically trained executive. I also directly estimate the difference in response of hospitals with and without a clinically trained executive:

    An important distinction is whether the classification of nonprofits as those with or without clinically trained executives is time-varying. For overall ownership status, switching is rare, and I limit to those who never switch. However, hospital executive teams are dynamic, and whether or not a hospital employs a clinically trained executive in each period may change. Therefore, I first present effects of having a clinically trained executive among stable executive teams, i.e., hospitals with no change in their propensity to hire a clinically trained executive.

@@

     First, I consider the differential program response of for-profit firms compared to the different sub-samples of nonprofit firms. These results are presented in Table \ref{tab:forprofit_readmort_synth}. Models in the left panel estimate the effect of being for-profit after the programs were enacted, compared to nonprofit sub-samples, on readmission rates. There is no difference between the response of for-profits and nonprofits as a whole, or nonprofits without a clinically trained executive. However, for-profits lower their readmissions by .3ppts compared to nonprofits without clinically trained executives. Models in the right panel estimate the effect of being for-profit after the programs on mortality rates. Here, there is no statistical difference in mortality rate response between different hospital types.

     \begin{figure}
         \centering
         \caption{Readmission Results}
         \includegraphics[width=\textwidth]{Objects/read_synth_plot.pdf}
         \label{fig:read_synth_plot}
     \end{figure}

     Next, I directly estimate the difference in response by nonprofits with and without clinically trained executives. These results are presented in Table \ref{tab:MD_noMD_readmort_synth}, where the left has models with outcome readmission rates and the right has models with outcome mortality rates. Nonprofits without clinically trained executives have a larger decrease in readmissions than those with clinically trained executive by .3ppts, while there is no difference in change in mortality rates. Tables \ref{tab:forprofit_readmort_synth} and \ref{tab:MD_noMD_readmort_synth} taken together show that, in terms of readmission rates, nonprofits without clinically trained executives behave more similarly to for-profits than they do other nonprofits. Under the assumption that having a clinically trained executive on the leadership team indicates a higher preference for social utility, these findings are consistent with the theoretical prediction presented in Section \ref{sec:model}.

>>>>

\usepackage{tgtermes}
\usepackage{epsf}
\usepackage{epstopdf}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage[colorlinks=true,linkcolor=blue,citecolor=blue]{hyperref}
\usepackage{dcolumn}
\usepackage{amsmath, amsthm, amssymb}
\usepackage{mwe}
\usepackage{url}
%\usepackage{harvard}
\usepackage{fancyheadings}
\usepackage{longtable}
\usepackage{authblk}
\usepackage{setspace}
%\usepackage[nomarkers]{endfloat}
\usepackage{float}
\usepackage{bbm}
%\usepackage{titling}
\usepackage{subcaption}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{import}
\usepackage[backend=biber,style=authoryear,
sorting=ynt,citestyle=authoryear]{biblatex}
\addbibresource{papercitations.bib}
%\usepackage[nomarkers,nofiglist,notablist]{endfloat}
\usepackage{subcaption}
\usepackage{caption}

@@

    \begin{abstract}
		{\small
        In many industries, not-for-profits (NFP) make up the majority of firms in the market. Understanding the objectives and behaviors of these firms is important when changing incentives. However, NFP firms have goals and missions that make understanding their objectives difficult. For hospitals in particular, there is mixed evidence on whether NFP hospitals differ meaningfully from for-profit (FP) hospitals. In this paper, I investigate whether leadership team composition of not-for-profit hospitals affects how similarly they behave to for-profits. I use a novel data set on NFP hospital executives in the US (2009-2015) to investigate whether the presence of clinically trained executives affects hospital behavior. In particular, I leverage two pay-for-performance initiatives in the US enacted in 2012, and estimate the difference in readmission and mortality rates afterwards between FP, NFP, and NFP with and without clinically trained executives. I find that NFPs without clinically trained executives behave more like FPs than other NFPs. This is consistent with theoretical predictions under the assumption that NFP hospitals with clinically trained executives value societal outcomes more than those without clinically trained executives.
		} 
	\end{abstract}

@@

    This paper contributes to three strands of literature. First, it adds to our understanding of what underlies NFP hospital behavior. There is mixed evidence on whether NFP behavior differs from FP behavior: costs, uncompensated care, technology adoption, and quality (\cite{sloan2000not}; \cite{eggleston2008hospital}; \cite{moscelli2018effect}; \cite{moscone2020public}). Underlying hospital are hospital objectives. To this point, researchers have found that NFPs do not act as purely profit maximizers, but maximize some combination of profit and output (\cite{deneffe2002not}, \cite{chang2011not-for-profit}). A paper with an approach similar to mine which investigates hospital behavior is \citeauthor{chang2011not-for-profit} (\citeyear{chang2011not-for-profit}), which looks at the effect of a cost shock on likelihood of shutting down and mix of profitable services. They find that NFPs and FPs are equally as likely to shut down after the shock, but only NFPs adjust their mix of profitable services. Similarly, I investigate a change in incentives to hospitals, but I look at the differential impacts of leadership teams within not-for-profits. This paper adds to our understanding of inputs into not-for-profit hospital objective functions.

@@

    \subsection{Not-for-Profit Hospital Leadership}

    Not-for-profit firms differ from for-profits mainly in that for-profit firms distribute revenue to stakeholders, while not-for-profits reinvest profit back into the firm. Another important distinction is that NFPs are, in general, driven by a mission to better society. Additionally, NFPs in the US receive tax benefits for this designation.

    NFPs are governed by a board of directors, whose role is to set broad goals and strategies, and provide general oversight. The board selects executives, the highest level of management of a firm, to carry out day-to-day operations. An executive team usually consists of at least a CEO and CFO, but every NFP is different in exactly how they structure their executive team. Many hospital executives specialize in health care settings by getting degree(s) in health care management, or an MBA specific to health care. There are specific positions that are often filled by someone with clinical training, such as a Chief Medical Officer, and doctors can even find themselves in other top managerial positions as well, such as CEO or president. While some doctors earn additional degrees before stepping into an executive role, this is not a necessary condition to becoming a physician executive.

    Anecdotal evidence is mixed on the benefits of hiring physician executives. On one hand, physicians bring a unique combination of clinical expertise and being comfortable with administration, and they can bring insights that improve hospital outcomes (\cite{Stajduhar_2023}, \cite{Ahmed_2022}). Alternatively, physicians may not always be adequately trained to enter into high level management positions, and therefore are not equipped to lead a firm (\cite{HarvardBusinessReview2018}).

    All tax-exempt organizations in the US are required to file a Form 990 with Internal Revenue Services (IRS) each year. There are different types of forms, but any organization grossing over \$200,000 must file the most extensive Form 990. Sections of this form include a statement of revenue, statement of functional expenses, a balance sheet, and, as used in this project, a list of all key employees, executives, and board members. Each firm is required to report the name and title, average hours per week, position, and compensation of their board members and executive leaders.

    \subsection{Pay-for-Performance Policies}\label{sec:hrrp}

    Two programs were passed as part of the Affordable Care Act that focused on pay-for-performance incentives for hospitals, the Hospital Readmissions Reduction Program (HRRP), and the Hospital Value Based Purchasing Program (HVBP). The HRRP focused on penalizing hospitals with poor quality in terms of readmission rates, and the HVBP fpcused on rewarding hospitals with high quality in terms of something.

    In October 2011, the Center for Medicare and Medicaid Services (CMS) released a set of rules under HRRP that mandated penalties for hospitals with above average readmission rates. A readmission is when a patient returns to the hospital within 30 days of being discharged from a previous stay; avoidable readmissions are a bad outcome for patients and increase health care spending. The goal of HRRP is to lower readmissions through better care coordination, less initial stay complications, and better post-care instructions. Beginning in October 2012, hospitals with higher readmission rates than the national average in pneumonia, heart failure, or AMI (after adjusting for demographic characteristics) receive a fixed lower reimbursement rate for all Medicare patients seen in their hospital. In 2015, CMS also included chronic obstructive pulmonary disease, coronary artery bypass graft surgery, and elective primary total hip arthroplasty and/or total knee arthroplasty as conditions which go into the penalty calculation (\cite{CMS}).

    Penalized hospitals pay penalties in the form of a fixed rate for every Medicare patient regardless of the condition. The rate began at 1\% and increased to 3\% in 2015. Further, CMS does not distinguish a necessary readmission from an avoidable readmission; any repeat hospital visit is included in the penalty calculation. Excess readmission rates are calculated using a rolling look-back period of 3 years to determine whether the hospital is penalized. Therefore, hospitals had incentive to react immediately once details of the program were announced in October of 2011. These penalties are not insignificant; penalized hospitals paid, on average, 4-5\% of revenue.

    The HVBP Program instead rewards hospitals with high quality or shown improvement in quality. Specifically, CMS deducts Medicare payments by 2\%, collects this sum, and divides it among the rewarded hospitals. Several quality and cost measures in categories safety, efficiency and cost reductions, clinical outcomes, and community engagement, are combined to create a single score metric for each hospital. Hospitals are then compared to each other and score points for being above average quality and for showing improvement (\cite{CMS_2023}).

@@

    One reason that hospital executive teams are understudied is because of the inaccessibility of the information. I construct a novel data set of not-for-profit hospitals in the US that contains names, titles, and positions of all executives tied to the hospital from 2009-2015. I gather this data from publicly available Tax Form 990s, which every sufficiently large not-for-profit files each year in the US. To my knowledge, this is the first large scale gathering of executive names from these forms.\footnote{\citeauthor{brickley2010board} (\citeyear{brickley2010board}) collects compensation information for a select number of hospitals.}

    The historical tax forms for all not-for-profits are housed by ProPublica.\footnote{https://projects.propublica.org/not-for-profits/} I use the not-for-profit Explorer API to extract Employee Identification Numbers (EIN) of not-for-profits classified as a hospital according to their National Taxonomy of Exempt Entities (NTEE) code. After filtering out associations and specialty hospitals, there are approximately 3,000 not-for-profit EINs. Importantly, linking EINs to other hospital information is crucial for the analysis. I rely on matching by location and name of hospital in the tax forms and AHA data. I confidently match 1200 EINs to an AHA identification number based on exact name matches within the same state. I compare the matched hospitals to non-matched hospitals in Appendix \ref{app:matched}, and find that the matched sample is not qualitatively different than the non-matched hospitals.

    For each of the 1200 EINs, I then extract web URLs to Form 990 PDFs in each year. I download these locally and use optical character recognition (OCR) text extraction methods to scrape the relevant section of the PDF, which is titled ``Officers, Directors, Trustees, Key Employees, and Highest Compensated Employees". Finally, I use string cleaning methods to identify names and positions for each EIN, year. I also identify titles indicating clinical training: MD, Dr., or DO. Using OCR extraction can be unreliable in some instances if the text is handwritten or small. Therefore, I manually record names, positions, and titles for observations that are missing after the initial extraction. This yields 852 not-for-profit hospitals that are matched to AHA data, and are not missing text information.

    Due to the lack of consistency in organization of executive teams, I focus mainly on the executive teams as a whole instead of specific positions. I drop anyone who is solely a board member, and focus on those who have executive in their title or are labeled as a president or vice president of a department. I create various hospital-level characterizations based on their executive teams, such as the existence of a clinically trained executive, the number of clinically trained executives, the number of total executives, and whether the hospital has a Chief Medical Officer.

    I merge the created sample of not-for-profits with all for-profits in the AHA survey. The main outcomes I consider in this paper are readmission and mortality rates, which are directly targeted in the pay-for-performance policies, and are from the CMS Hospital Compare data. I combine readmission and mortality rates for pneumonia, AMI, and heart failure (the relevant HRRP conditions) into a weighted average based on the number of patients in each condition. I also include as outcomes uncompensated care and case mix index (a measure of patient complexity). While not directly targeted by the policies, uncompensated care is a mechanism by which a hospital might re-allocate resources by taking on less charitable cases, and case mix index reflects potential gamesmanship in the type of patients admitted to the hospital. Uncompensated care comes from the Healthcare Cost Information Reporting Systems (HCRIS) data, and case mix index is from CMS public use files.

    Additionally, I include several relevant hospital characteristics. Beds, whether the hospital is an academic medical center, owns subsidiaries, is physician owned or system affiliated, come from the AHA survey. Whether the hospital was penalized under HRRP or received payments under HVBP is found in the HCRIS data. Finally, I limit to years 2010-2014.

    % In Table \ref{sumstats}, I present summary statistics for hospital variables. There are 3,766 total hospitals in the sample, with 150 beds on average. Hospitals in the sample are less likely to be penalized for AMI than either heart failure or pneumonia under the HRRP program. On average, 20\% of patients with one of these conditions would be readmitted. While readmission rates for the specific conditions are similar, the highest likelihood of readmission is for heart failure patients. Alternatively, AMI is the most dangerous condition in terms of mortality rates.

    I present a table of means for each hospital type (for-profit, in-sample not-for-profits, and not-for-profits with and without a clinically trained executive at some point in time) in Table \ref{tab:sumstats_samples}. Half of the hospitals in the sample are classified as not-for-profit with leadership information. There are key differences in not-for-profits with and without clinically trained executives: those with MDs are larger, more likely to be penalized, have higher uncompensated care, and more complex patients. I discuss the implications of these differences for identification in Section \ref{sec:model}.

@@

    I plot readmission and mortality rates across time for different hospital types in Figure \ref{fig:weighted_read_mort_graph}. For-profit hospitals have the highest readmission rates prior to 2012. Not-for-profits with MDs have readmission rates that fall between for-profits and not-for-profits without MDs. While the trend in readmission rates behaves similarly across hospital types prior to the programs taking affect, there are differential trends after the programs. NFPs with an MD have the lowest mortality rates before the programs, but differentially continue with an increasing trend in mortality after the programs. While this is purely descriptive evidence, the trends in outcomes point to more similar behavior between for-profits and NFPs without clinically trained executives in comparison to NFPs with a clinically trained executive.

    \begin{figure}[ht!]
    \centering
        \caption{Outcomes Over Time}
        \includegraphics[width=\textwidth]{Objects/weighted_read_mort_graph.pdf}
        \label{fig:weighted_read_mort_graph}
    \end{figure}

    I model hospital behavior in two worlds: one where quality does not directly affect profit, and one where it does. Pay-for-performance incentives essentially incorporate quality of care into a hospital's profit function. Intuitively, a hospital that places more weight on profit (for-profit type) will only care about quality when it enters into the profit function, and therefore will respond more drastically than a hospital who already cared about quality before the incentives. Therefore, hospital response to pay-for-performance incentives reveals something about their underlying preference on profit vs. societal outcomes.

    Formally, I model the hospital objective function as a weighted average of profit and societal (or patient) outcomes. For simplicity, I abstract away from quality having a direct effect of quantity and price, though these are possible and don't change the important findings here. Without any financial incentives to increase quality, this looks like:

    \noindent where $R$ is net revenue, $c_{\theta}(.)$ is an increasing cost function specific to quality $\theta$, $u(.)$ is extra utility gained from quality, which is increasing and concave in $\theta$, and $\alpha\in[0,1]$ captures how much the hospital cares about profit vs. societal benefit. There is an implicit stay open condition throughout the model. Taking the first order condition yields

    $$\frac{du'(\theta)}{d\alpha} = \frac{1}{(1-\alpha)^2}c_{\theta}'(\theta) > 0.$$

    \noindent That is, the more weight on profit, the lower quality of care they will provide.

    In a world with pay-for-performance incentives, which can either look like benefits from high quality (such as HVBP) or penalties for low quality (as in HRRP), quality directly affects revenue. Thus, $R(\theta)$ is an increasing function of $\theta$, and I assume that the marginal financial benefit of increasing quality is greater than the marginal cost ($R'(\theta)\geq c_{\theta}'(\theta))$. Thus, the new first order condition yields:

    $$u'(\theta) = \frac{\alpha}{1-\alpha} \left[c_{\theta}'(\theta)-R(\theta)\right] \leq 0.$$

    \noindent Taking the derivative of this with respect to $\alpha$,

    $$\frac{du'(\theta)}{d\alpha} = \frac{1}{(1-\alpha)^2}c_{\theta}'(\theta).$$

    Thus,

    $$\frac{d\theta}{d\alpha} = \frac{du'(\theta)/d\alpha}{du'(\theta)/d\theta}\geq0.$$

    Therefore, in a world with financial incentives on quality, quality is increasing with more weight placed on profit. The purpose of this paper is to investigate the response of different types of hospitals when moving from a world closer to this stylized world without pay-for-performance to the stylized world with pay-for-performance. Hence, I combine these results into one response function that depends on $\alpha$, where $\theta_2$ is quality under pay-for-performance incentives and $\theta_1$ is quality with no financial incentive to quality:

    \begin{align*}
        \frac{d\Delta\theta}{d\alpha}&=\frac{d(\theta_2-\theta_1)}{d\alpha}\\
        &=\frac{d\theta_2}{d\alpha}-\frac{d\theta_1}{d\alpha}\\
        &\geq 0.
    \end{align*}

    The theoretical prediction is that change in quality depends on how much weight the hospital places on profit vs. societal benefit. Particularly, hospitals with more weight on profit respond more than hospitals with more weight placed on societal benefit. In reality, $\alpha$ is unknown for many hospitals. What is observed is ownership status, where for-profit hospitals have revealed that they have a high $\alpha$. However, not-for-profit behavior on average does not bring much clarity about their objectives (see Section for the literature review). Thus, I compare the response to incentive change between for-profits and not-for-profits with different leadership teams to bring clarity on how these things point to an underlying $\alpha$.

    In light of the theoretical predictions, I first test empirically whether different sub-samples of not-for-profits respond to program incentives differently than for-profits. I estimate the following regression equation:

    \noindent where $y_{ht}$ is one of the outcome variables discussed in Section \ref{sec:data}, $\beta$ is the coefficient of interest capturing the effect of being for-profit after the programs were introduced, and $\gamma_h$ and $\delta_t$ are hospital and time fixed effects, respectively. For each outcome, I estimate three models where the comparison group differs by hospital type: all NFPs, NFPs with a clinically trained executive, and NFPs without a clinically trained executive. I also directly estimate the difference in response of NFPs with and without a clinically trained executive:

    An important distinction is whether the classification of NFPs by leadership team is time-varying. For overall ownership status, switching is rare, and I limit to those who never switch. However, hospital executive teams are dynamic, and whether or not a hospital employs a clinically trained executive in each period may change. Therefore, I first present effects of having a clinically trained executive among stable executive teams, i.e., NFPs with no change in their propensity to hire a clinically trained executive.

    \subsection{Identification}

    While ultimately I employ a synthetic difference-in-differences method, the conceptual identification is identical to a standard difference-in-differences.

    \subsection{Decomposition}

@@

@@

     I consider the differential program response in readmission rates of for-profit firms compared to the different sub-samples of not-for-profit firms (specification \ref{eq:forprofit}). First, I compare for-profits to all not-for-profits in the sample in Figure \ref{fig:read_synth_plota}, which shows that on average, there is no difference in how for-profits and not-for-profits change their readmission rates in response to the incentive changes. Figures \ref{fig:read_synth_plota} and \ref{fig:read_synth_plotb} compare for-profit readmission rates to NFPs with a clinically trained executive and without a clinically trained executive, respectively. Figure \ref{fig:read_synth_plota} shows that for-profit hospitals decrease readmission rates more than not-for-profits with an MD executive after pay-for-performance incentives. However, while for-profits have higher readmission rates than NFPs without a clinically trained executive, there is no difference in their change in readmission rates after the incentives are enacted, shown in Figure \ref{fig:read_synth_plotb}.
